[Exercise Set 6](exercises/Exercise6.html).

### Data Wrangling

We prepare BPRS and RATS datasets in `meet_and_repeat.R` and save them in `data` folder.

`BPRS` contains 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS is used to evaluate patients suspected of having schizophrenia.

`RATS` data is from a nutrition study conducted in three groups of rats. The groups were put on different diets, and each animal’s body weight (grams) was recorded over a 9-week period (two times in week 7). The question of most interest is whether the growth profiles of the three groups differ.

```{r}
# run meet_and_repeat.R to initialize both datasets
source("data/meet_and_repeat.R", local = knitr::knit_global())
```

------------------------------------------------------------------------

### RATS Analysis

First, let's plot the rat weight trendlines for each group:

```{r}
# access libraries
library(tidyr); library(dplyr); library(ggplot2); library(readr);

# Plotting data
ggplot(RATS, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() + 
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "none")
```

The initial weights differ between groups, so we should also look at standardized values.

```{r}
# mutate to add stdweights column 
RATS <- RATS %>%
  group_by(Time) %>%
  mutate(StdW = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

ggplot(RATS, aes(x = Time, y = StdW, linetype = ID)) +
  geom_line() + 
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (standardized)") +
  theme(legend.position = "none")
```

Rats with larger initial weight tend to gain weight faster. After standardization this is not necessarily so.

We can also create a summary graph for each group, by adding mean and standard error columns to the dataset, and plotting the trends with errorbars:

```{r}
# summary dataset
RATSS <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(length(Weight)) ) %>%
  ungroup()

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = 'right') +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Repeating with stadardized weights:

```{r}
# summary dataset
RATSS <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(StdW), se = sd(StdW)/sqrt(length(StdW)) ) %>%
  ungroup()

ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = 'right') +
  scale_y_continuous(name = "mean(StdW) +/- se(StdW)")
```
Boxplots of weights for each group:

```{r}
RATSS <- RATS %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight) ) %>%
  ungroup()

ggplot(RATSS, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Week)")
```
Boxplots don't have massive outliers (although one rat in group 2 could be considered one). 

Next, we apply a t-test to assess any difference between the groups, and also calculate a confidence interval for this difference. There are 3 groups so I performed the t-test for groups 2 and 3.

```{r}
RATS2 <- RATS %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Perform a two-sample t-test
t.test(mean ~ Group, data = filter(RATS2, RATS2$Group != 1), var.equal = TRUE)
```
There is no significant evidence for difference between groups. Also the 95% confidence interval is wide and includes the zero, allowing for similar conclusions to be made.

We can also fit a linear model and compute the analysis of variance (ANOVA) for the model.

```{r}
# Add the baseline from the original data as a new variable to the summary data
RATS1 <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep ="\t", header = T)
RATS2 <- RATS2 %>% mutate(baseline = RATS1$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATS2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

P-value is not significant for between groups.

------------------------------------------------------------------------

### BPRS Analysis

Plotting BPRS per treatment group:

```{r}
ggplot(BPRS, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)))
```

First we can fit a multiple linear regression model:

```{r}
# create a regression model
fit <- lm(bprs ~ week + treatment, data = BPRS)
# print out a summary of the model
summary(fit)
```

There is a significant p-value difference over time, but not with different treatments.

The previous model assumes independence of the repeated measures of weight, and this assumption is highly unlikely. So, now we will move on to consider both some more appropriate graphics and appropriate models. We will first fit the random intercept model for the same two explanatory variables using lme4 package. Fitting a random intercept model allows the linear regression fit for each subject to differ in intercept from other subjects.

```{r}
# access library lme4
library(lme4)
# Create a random intercept model
fit <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)
# Print the summary of the model
summary(fit)

```

Now we can move on to fit the random intercept and random slope model `(week | subject)` to the data. Fitting a random intercept and random slope model allows the linear regression fits for each individual to differ in intercept but also in slope. This way it is possible to account for the individual differences in the subject's profiles, but also the effect of time. We also allow for a `week` * `treatment` interaction.

```{r}
fit <- lmer(bprs ~ week + treatment + (week | subject) + week*treatment, data = BPRS, REML = FALSE)
summary(fit)
```

Plotting the fitted we can see that the model did not find difference between the two groups:

```{r}
# Create a vector of the fitted values
Fitted <- fitted(fit)

# Create a new column fitted
BPRS <- BPRS %>% mutate(Fitted = Fitted)

ggplot(BPRS, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$bprs), max(BPRS$bprs)))

```



